---
title: "11 Confounding and interaction in regression"
subtitle: Applied regression analysis and other multivariate methods
author: "[Yi Zhou](https://github.com/zy0426/Book-Applied-regression-analysis)"
date: "February 24, 2016"
fontsize: 10pt
output: 
  beamer_presentation:
    fig_height: 4
    fig_width: 4
---

## Preview

Two goals in a regression analysis

- To predict the dependent variable (Y), to find a model
- To quantify the relationship between X and Y, to produce statistical inference about coefficients

The second goal is of paticular interest in clarufying a causal process

## Overview

- Interaction

    - The relationship of interest is different at different levels of the extraneous variables
    - e.g., dose the `PAL`-`SBP` relationship vary with `AGE`
    - Require a statistical test
    
- Confounding

    - Meaningfully different interpretations of the relationship of interest result when the extraneous variable is ingnored or included
    - e.g., dose the `PAL`-`SBP` relationship vary if we ignore `AGE`
    - Comparison between a crude estimate of an association and an adjusted one
    - Do not require a statistical test
    

## Interaction

Case 1

```{r, echo=FALSE}

data1 <- data.frame(
  T = c(1,2, 1,2),
  C = c(1,1, 2,2),
  Y = c(1,3, 2,4)
)
l1 <- lm(Y~T, data1[1:2, ])
l2 <- lm(Y~T, data1[3:4, ])

plot(data1$T, data1$Y)
abline(l1)
abline(l2, col = "red")
```

##


```{r, echo=FALSE,results='asis'}
stargazer::stargazer(l1,l2,  header = FALSE, column.sep.width = "2pt",
                     no.space = TRUE, font.size ="small", single.row=TRUE,
                     column.labels = c("black", "red"),
                     dep.var.labels.include = FALSE,
                     dep.var.caption = "Y",
                     digits = 0)
```



## 

Case 2

```{r, echo=FALSE}
data2 <- data.frame(
  T = c(1,2, 1,2),
  C = c(1,1, 2,2),
  Y = c(1,3, 4,2)
)
l1 <- lm(Y~T, data2[1:2, ])
l2 <- lm(Y~T, data2[3:4, ])
l3 <- lm(Y~T*C, data2)

plot(data2$T, data2$Y)
abline(l1)
abline(l2, col = "red")
```

##


```{r, echo=FALSE,results='asis'}
stargazer::stargazer(l1,l2,l3,  header = FALSE, column.sep.width = "2pt",
                     no.space = TRUE, font.size ="small", single.row=TRUE,
                     column.labels = c("black", "red", "total"),
                     dep.var.labels.include = FALSE,
                     dep.var.caption = "Y",
                     digits = 0)
```

## Interaction modeling in general

Three approaches to spicify which terms to include

- interactions resonable a prior based on literature and expertise
- a full set of product terms
- interaction with the primary factors

## Effect modification 

Effect-measurement modification

- Variation in the magnitude of a measure of exposure effect across levels of another variable
- A finding to be reported
- `Statistics` and `epidemiology`
- heterogeniety of effect, nonuniformity of effect, effect variation
- risk-difference modification vs. risk-ratio modification
- the presence/abscence of (statistical) interaction is decided by the (epidemiological) scale (difference/ratio) 

## Confounding

We have to determine subjectively

- A bias to be avoided
- Confounding should take precedence over precision
- Interactino should take precedence over confounding

